---
title: "Introduction to the cellKey-Package"
author: "Bernhard Meindl"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
toc_depth: 5
number_sections: false
vignette: >
  %\VignetteIndexEntry{Introduction to the cellKey-Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

  ```{r, echo=FALSE}
library(rmarkdown)
```

# Introduction

TBW

Implementation based on paper *Methodology for the Automatic Confidentialisation of Statistical Outputs from Remote Servers at the Australian Bureau of Statistics* (Thompson, Broadfoot, Elazar).

## Main Features

TBW

- using sha1-sums on input objects to set seed for record-key generation if not possible
- record-keys can be pre-generated

## Todo
- ~~allow hierarchical input in  `perturbTable()`~~
- ~~check parametrisation of perturbation tables~~
- performance optimisations

## An Example

### Load the Package
```{r}
library(cellKey)
packageVersion("cellKey")
```

### Use Testdata
```{r}
dat <- ck_create_testdata()
dat$age <- as.integer(cut(dat$age, 6))
head(dat)
```

### create required input
```{r}
pert_params <- ck_create_pert_params(
  bigN=17312941, 
  smallN=12,
  pTable=ck_create_pTable(pTableSize=70),
  sTable=ck_generate_sTable(smallC=12), 
  mTable=c(0.6,0.4,0.2))
inp <- ck_create_input(dat=dat, def_rkey=15*nrow(dat), pert_params=pert_params)
```

### create dimensions
```{r}
dim.sex <- data.table(levels=c("@","@@","@@"), codes=c("Total", 1, 2))
dim.age <- data.table(levels=c("@",rep("@@", 6)), codes=c("Total", 1:6))
dimList <- list(sex=dim.sex, age=dim.age)
```

These dimensions must be specified the same way is in **sdcTable**. 

```{r}
print(dimList)
```


### Compute perturbed tables

- unweighted and only counts

```{r}
tab1 <- perturbTable(inp, dimList=dimList, weightVar=NULL, numVars=NULL)
results(tab1)
```

- weighted and also numerical variables

```{r}
tab2 <- perturbTable(inp, dimList=dimList, weightVar="sampling_weight", numVars=c("savings","income"))
results(tab2, meanBeforeSum=TRUE)
```

### extract information from results

```{r}
slotNames(tab2)

# information on perturbations for counts
mod_counts(tab2) 

# information on perturbations for numerical variables
mod_numvars(tab2) 
```


